TBD

### 2018-10-19 paper/paper.tex
new sections
filled in some comments

\section{introduction}\label{sec:intro}
	Gaussian processes can be used to model correlated variability in astrophysical time-series. They have been used to study stellar variability, to analyze AGN 
	variability, and to model correlated noise in photometry and radial velocity observations of exoplanet host stars \gordontodo{more applications outside of exoplanets? 
	X-ray binaries, modeling dust distributions, etc...}.

	Gaussian process noise models are most 
	advantageous in cases where correlated noise dominates over white noise. This is already the relevant case for state-of-the-art radial velocity observations, but transit 
	photometry is not far behind. High-precision photometry from future space-based observatories such as James Webb will enable detection of ever more shallow 
	\gordontodo{shallower?}
	transit signals, so long as they can be distinguished from correlated variability. \gordontodo{This is where we come in...}

@@

	\gordontodo{What about other methods of accounting for astrophysical noise? i.e. cosine filtering, polynomials, etc.}

@@

\section{2d gaussian processes}
	\subsection{kronecker structure of the 2d covariance matrix}
	In two dimensions the kernel function depends on two coordinates. We will continue to use
	\begin{equation}
		\bvec{t} = \expandvec{t_1}{t_N}
	\end{equation}
	to refer to the first coordinate, and introduce
	\begin{equation}
		\bvec{u} = \expandvec{u_1}{u_M}
	\end{equation}
	to refer to the second coordinate. In most cases, we will demand that $M << N$. We then define $\bvec{x}$ to be the list of points
	\begin{equation}
		\bvec{x} = \expandvec{(t_1, u_1), (t_1, u_2), \cdots (t_1, u_M)}{(t_N, u_M)}
	\end{equation}
	We can also write this as
	\begin{equation}
		x_i = (t_{\lfloor(i-1)/M\rfloor + 1}, u_{i - \lfloor(i-1)/M\rfloor M})
	\end{equation} 
	The covariance between $x_i$ and $x_j$ is then given by
	\begin{equation}
		k_{\bvec{\alpha}\bvec{\beta}}(x_i, x_j) = \sigma_i^2\delta_{ij} + q_\bvec{\alpha}(\tau_{ij})\circ r_\bvec{\beta}(\nu_{ij})
	\end{equation}
	where $\tau_{ij} = |t_{\lfloor(i-1)/M\rfloor + 1}-t_{\lfloor(j-1)/M\rfloor + 1}|$, 
	$\nu_{ij} = |u_{i - \lfloor(i-1)/M\rfloor M}-u_{j - \lfloor(j-1)/M\rfloor M}|$; 
	$q$ and $r$ are one-dimensional kernel functions which depend on the hyperparameters $\bvec{\alpha}$ and 
	$\bvec{\beta}$ respectively; 
	and $\circ$ is some binary operation. 
	If $Q_\bvec{\alpha}$ and $R_\bvec{\alpha}$ are the covariance matrices corresponding to 
	$q_\bvec{\alpha}$ and $r_\bvec{\alpha}$ such that
	\begin{equation}
		\left[Q_\bvec{\alpha}\right]_{ij} = q_\bvec{\alpha}(\tau_{ij})\ \ \ \mathrm{and}\ \ \ \left[R_\bvec{\beta}\right]_{ij} = r_\bvec{\beta}(\tau_{ij})
	\end{equation} 
	then, using $q_{ij}$ and $r_{ij}$ as shorthand for the entries in $Q_\bvec{\alpha}$ and $R_\bvec{\beta}$, $K_\bvec{\alpha\beta}$ can be represented as: 
	\begin{equation}
		K_\bvec{\alpha\beta} = \Sigma + 
			\begin{bmatrix}
				q_{11}\circ R_\bvec{\beta} & ... & q_{1N}\circ R_\bvec{\beta} \\
				\vdots & \ddots & \\
				q_{N1}\circ R_\bvec{\beta} & ... & q_{NN}\circ R_\bvec{\beta}
			\end{bmatrix}
	\end{equation}
	where $[\Sigma]_{ij} = \sigma_i^2\delta_{ij}$ and
	\begin{equation}
		q_{pq}\circ R_\bvec{\alpha} = 
			\begin{bmatrix}
				q_{pq}\circ r_{11} & ... & q_{pq}\circ r_{1M} \\
				\vdots & \ddots & \\
				q_{pq}\circ r_{M1} & ... & q_{pq}\circ r_{MM}
			\end{bmatrix}
	\end{equation}
	If the operator $\circ$ represents multiplication, we have 
	\begin{equation}
		K_\bvec{\alpha\beta} = \Sigma + Q_\bvec{\alpha} \otimes  R_\bvec{\beta}
	\end{equation}
	where $\otimes$ is the Kronecker product. If the covariance in $\bvec{t}$ is described by a \celerite kernel can write 
	\begin{eqnarray}
		K &=& \Sigma + \left[\mathrm{tril}(UV^\T) + \mathrm{triu}(VU^\T)\right]\otimes R \\
		&=& \Sigma + \mathrm{tril}\left[ (U\otimes R)(V\otimes I_M)^\T\right] + \mathrm{triu}\left[ (V\otimes I_M)(U\otimes R)^\T\right]
	\end{eqnarray}
	This covariance has the same semi-separable form as in the one-dimensional case, but now $K$ is an $(NM\times NM)$ matrix of rank $2JM$. 
	This means that we can apply the same Cholesky decomposition algorithm (described in the appendix) as in the 1d case to compute the GP in 
	$\mathcal{O}(NJ^2M^3)$.

	\subsection{performance}
	Computation of a \celerite GP takes $\mathcal{O}(NJ)$ where $N$ is the length of the coordinate vector $\bvec{x}$ and $J$ is the number of 
	\celerite terms making up the kernel function.

\section{example: exomoons}

>>>>

\section{introduction}\label{sec:intro}
	Gaussian processes have been used extensively in astrophysics to model stochastic variability across space and time. 
	They have been used to study variable stars and AGN, to model the extragalactic distribution of dust, and to model correlated noise in 
	photometric and radial velocity observations of exoplanet host stars.	\gordontodo{X-ray binaries?}

	\gordontodo{The following paragraph might make more sense in the exomoons section.}
	Gaussian process noise models are most advantageous when correlated noise dominates over white noise. 
	This is already the case for state-of-the-art radial velocity observations, and transit 
	photometry is not far behind. Future space-based observatories such as James Webb 
	will enable high-precision photometry in which stellar variability is the main barrier to detecting ever more shallow transit signals. 
	Stellar variability which obscures transit signal can be a result of spot and faculae crossings, asteroseismic oscillations, or variable 
	granulation on the star's surface. GPs are able to \gordontodo{something about how a GP model is non-parametric and the advantages thereof? Also agnostic 
	to the source of variability (seems related)}

	\gordontodo{This is where we come in...}

@@

	\gordontodo{What about other methods of accounting for astrophysical noise? i.e. cosine filtering, polynomials, etc. --- also related to GP as non-parametric 
	noise model}

@@

\section{2d gaussian processes}
	\subsection{kronecker structure of the 2d covariance matrix}
	In two dimensions the kernel function depends on two coordinates. We will continue to use
	\begin{equation}
		\bvec{t} = \expandvec{t_1}{t_N}
	\end{equation}
	to refer to the first coordinate, and introduce
	\begin{equation}
		\bvec{u} = \expandvec{u_1}{u_M}
	\end{equation}
	to refer to the second coordinate. In most cases, we will demand that $M << N$. We then define $\bvec{x}$ to be the list of points
	\begin{equation}
		\bvec{x} = \expandvec{(t_1, u_1), (t_1, u_2), \cdots (t_1, u_M)}{(t_N, u_M)}
	\end{equation}
	We can also write this as
	\begin{equation}
		x_i = (t_{\lfloor(i-1)/M\rfloor + 1}, u_{i - \lfloor(i-1)/M\rfloor M})
	\end{equation} 
	The covariance between $x_i$ and $x_j$ is then given by
	\begin{equation}
		k_{\bvec{\alpha}\bvec{\beta}}(x_i, x_j) = \sigma_i^2\delta_{ij} + q_\bvec{\alpha}(\tau_{ij})\circ r_\bvec{\beta}(\nu_{ij})
	\end{equation}
	where $\tau_{ij} = |t_{\lfloor(i-1)/M\rfloor + 1}-t_{\lfloor(j-1)/M\rfloor + 1}|$, 
	$\nu_{ij} = |u_{i - \lfloor(i-1)/M\rfloor M}-u_{j - \lfloor(j-1)/M\rfloor M}|$; 
	$q$ and $r$ are one-dimensional kernel functions which depend on the hyperparameters $\bvec{\alpha}$ and 
	$\bvec{\beta}$ respectively; 
	and $\circ$ is some binary operation. 
	If $Q_\bvec{\alpha}$ and $R_\bvec{\alpha}$ are the covariance matrices corresponding to 
	$q_\bvec{\alpha}$ and $r_\bvec{\alpha}$ such that
	\begin{equation}
		\left[Q_\bvec{\alpha}\right]_{ij} = q_\bvec{\alpha}(\tau_{ij})\ \ \ \mathrm{and}\ \ \ \left[R_\bvec{\beta}\right]_{ij} = r_\bvec{\beta}(\tau_{ij})
	\end{equation} 
	then, using $q_{ij}$ and $r_{ij}$ as shorthand for the entries in $Q_\bvec{\alpha}$ and $R_\bvec{\beta}$, $K_\bvec{\alpha\beta}$ can be represented as: 
	\begin{equation}
		K_\bvec{\alpha\beta} = \Sigma + 
			\begin{bmatrix}
				q_{11}\circ R_\bvec{\beta} & ... & q_{1N}\circ R_\bvec{\beta} \\
				\vdots & \ddots & \\
				q_{N1}\circ R_\bvec{\beta} & ... & q_{NN}\circ R_\bvec{\beta}
			\end{bmatrix}
	\end{equation}
	where $[\Sigma]_{ij} = \sigma_i^2\delta_{ij}$ and
	\begin{equation}
		q_{pq}\circ R_\bvec{\alpha} = 
			\begin{bmatrix}
				q_{pq}\circ r_{11} & ... & q_{pq}\circ r_{1M} \\
				\vdots & \ddots & \\
				q_{pq}\circ r_{M1} & ... & q_{pq}\circ r_{MM}
			\end{bmatrix}
	\end{equation}
	If the operator $\circ$ represents multiplication, we have 
	\begin{equation}
		K_\bvec{\alpha\beta} = \Sigma + Q_\bvec{\alpha} \otimes  R_\bvec{\beta}
	\end{equation}
	where $\otimes$ is the Kronecker product. If the covariance in $\bvec{t}$ is described by a \celerite kernel can write 
	\begin{eqnarray}
		K &=& \Sigma + \left[\mathrm{tril}(UV^\T) + \mathrm{triu}(VU^\T)\right]\otimes R \\
		&=& \Sigma + \mathrm{tril}\left[ (U\otimes R)(V\otimes I_M)^\T\right] + \mathrm{triu}\left[ (V\otimes I_M)(U\otimes R)^\T\right]
	\end{eqnarray}
	This covariance has the same semi-separable form as in the one-dimensional case, but now $K$ is an $(NM\times NM)$ matrix of rank $2JM$. 
	This means that we can apply the same Cholesky decomposition algorithm (described in the appendix) as in the 1d case to compute the GP in 
	$\mathcal{O}(NJ^2M^3)$. While our implementation only handles the Kronecker product case, 
	his result holds in the case that $\circ$ is an operation other than multiplication, so long as $\circ$ has the property
	\begin{equation}
		(a\circ B)(c\circ D) = (ac)\circ(BD)
	\end{equation}
	for scalars $a$ and $b$ and matrices $B$ and $D$. 
	%where $a\circ B$ is the matrix 
	%\begin{equation}
	%	a\circ B = \begin{bmatrix}
	%		a\circ b_{11} & a\circ b_{12} & \cdots \\
	%		a\circ b_{21} &  a\circ b_{22} & \\
	%		\vdots & & \ddots & 
	%	\end{bmatrix}
	%\end{equation}
	\gordontodo{See appendix for proof}

	\subsection{sampling the distribution}
	A gaussian process can be used to simulate correlated noise. If $L$ is the lower triangular factor from the Cholesky decomposition of the 
	$(N\times N)$ covariance matrix $K$, and $\bvec{z}$ is a vector of length $N$ containing random numbers drawn for a standard normal distribution, i.e.
	\begin{equation}
		\bvec{z} \sim \mathcal{N}(1, 0)
	\end{equation}
	then $\bvec{x} = \bvec{\mu_x} + L\bvec{z}$ is a vector with entries correlated by $K$ and with mean $\bvec{\mu_x}$, i.e.
	\begin{equation}
		\bvec{x} \sim \mathcal{N}(\bvec{\mu_x}, K)
	\end{equation}
	\gordontodo{fix this so that $\mu$ depends on $\theta$ and is a function of $\bvec{t}$ or something like that. Compuational cost should be $\mathcal{O}(NJM^2)$}

	\subsection{interpolation and extrapolation}	
	\gordontodo{Bit more complicated --- if both dimensions are celerite processes then the scaling is better than in the case where $R$ is not celerite. You have notes on this.}

\section{example: exomoons}
	\gordontodo{What if I used Kepler-1625 as an example? Could simulate WFC3 GRISM data and show that moon could be validated with Celerite2D?}
